---
title: "Weighting behavior rating data to adjust for demographically non-representative samples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Problem Statement

To standardize and develop norms for behavior rating scales, we collect nationwide samples comprised of data from many thousands of persons. Our intent is to have the demographic makeup of these samples match the U.S. census proportions, as closely as possible. The demographic variables of interest are age, gender, (parent) education level, race/ethnicity, and U.S. geographic region.

Our approach is to stratify the sample by age, and ensure that we have enough cases within each age group to develop stable norms. Within each age stratum, we attempt to match the proportions of gender, education level, race/ethnicity, and region to the census proportions for that age group. Inevitably, some demographic cells are harder to fill than others, especially lower levels of education and certain ethnic minority groups. Pursuing the last few cases to fill out these cells is expensive and time-consuming.

We are interested in weighting the data of these samples that fall short of the ideal demographic composition. What I mean by this is to calculate weighting multipliers corresponding to each possible crossing of the four demographic variables of interest. For example, one crossing would be _female_ X _high-school diploma_ X _hispanic_ X _south region_. All cases falling into that particular bucket would have the same weighting multiplier.

That multiplier would be applied to the data at the most granular level. Consider the example of a behavior rating scale where a parent answers 50 questions (items) about their child's level of hyperactive/inattentive behavior during the past month. These behaviors are rated on a frequency scale, coded as follows:

* Never (or almost never) = 1
* Occasionally = 2
* Frequently = 3
* Always (or almost always) = 4

A higher code on any item indicates greater frequency of problematic behavior.

What I envision is that a demographic weighting multiplier would be applied case-wise to each of the 50 item codes. These data would then be weighted for all subsequent item- and score-level analyses.

In practice this would work as follows. Suppose that the previously mentioned demographic cell (_female_ X _high-school diploma_ X _hispanic_ X _south region_) was under-represented in our sample, with respect to U.S. census proportions. The demographic weighting multiplier would serve to increase the numerical impact of the insufficient number of cases in this cell. Thus, if we collected 40 cases in this cell, and the census required 60, the multiplier would be applied to increase the values of the item codes in our 40 cases. Once these values were increased, summing the item codes from from our 40 cases would yield a sum approximately equivalent to the sum of the item codes of 60 non-weighted cases with the same demographic characteristics. 

To recount, this is how I imagine we would use weighting to adjust the data of our collected sample, so that its statistics would be similar to those obtained from a sample that exactly matched the U.S. Census demographic proportions.

At this point I must pause to check my logic. Am I thinking about the problem and its solution in the right way? Is the application of case-wise demographic weights, in the manner that I’ve described, the correct method for adjusting the data of a sample that does not match census demographic targets?


### Work to date with Survey package

Here I will show you my work with the the Survey functions in a reproducible example.

#### Read unweighted input data

First we read in a simulated data set that I created with the `psych` package. The data set includes 1000 cases with the four demographic variables of interest, and 50 behavior rating items coded as described above. (Note: if you are unable to read the data set from the remote URL, I can provide code that will enable you to create the data.)
```{r read-data, eval=FALSE}
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(survey)))
set.seed(123)

urlRemote_path  <- "https://raw.github.com/"
github_path <- "wpspublish/DSHerzberg-WEIGHTING-DATA/master/INPUT-FILES/"
fileName_path   <- "unweighted-input.csv"

unweighted_input <- suppressMessages(read_csv(url(
  str_c(urlRemote_path, github_path, fileName_path)
)))
```

<br>

#### Compare demographic counts to census targets

Running the next code chunk yields a table `freq_demos_comp` that compares the demographic counts of the unweighted input to the target proportions of the U.S. Census (which are captured in the vectors with the suffix `_census`). `freq_demos_comp` is sorted by the right-most column `diff_pct`, which expresses the difference between the unweighted input and the census target, in the sample percentages for each demographic category.

The table shows, for example, that the `female-no_HS-hispanic-northeast` cell is the most under-represented cell in the input, whereas the `male-BA_plus-white-midwest` cell is the most over-represented. By contrast, the input sample is closer to the census proportions for blacks or asians of either gender from the `south` or `west`, with educational levels of `HS_grad` or  `some_college`.

By the logic expressed previously, I would therefore expect to require a demographic weighting multiplier greater than 1 for cases in the `female-no_HS_hispanic_northeast` cell, and a multiplier less than 1 for cases in the `male-BA_plus-white-midwest` cell.
```{r demo-counts, eval=FALSE}
var_order <- c("age", "age_range", "gender", "educ", "ethnic", "region", "clin_status")

cat_order <- c(
  # age
  NA, "5", "6", "7", "8", "9", "10", "11", "12",
  # age_range
  NA, "5 to 8 yo", "9 to 12 yo", 
  # Gender
  NA, "male", "female",
  # educ
  NA, "no_HS", "HS_grad", "some_college", "BA_plus",
  # Ethnicity
  NA, "hispanic", "asian", "black", "white", "other",
  # Region
  NA, "northeast", "south", "midwest", "west")

gender_census <- tibble(cat = c("female", "male"),
                        Freq = nrow(unweighted_input)*c(0.53, 0.47))
educ_census <- tibble(cat = c("no_HS", "HS_grad", "some_college", "BA_plus"),
                      Freq = nrow(unweighted_input)*c(0.119, 0.263, 0.306, 0.311))
ethnic_census <- tibble(cat = c("hispanic", "asian", "black", "white", "other"),
                        Freq = nrow(unweighted_input)*c(0.239, 0.048, 0.136, 0.521, .056))
region_census <- tibble(cat = c("northeast", "south", "midwest", "west"),
                        Freq = nrow(unweighted_input)*c(0.166, 0.383, 0.212, 0.238))

freq_demos_unweighted <- unweighted_input %>%
  pivot_longer(age_range:clin_status, names_to = 'var', values_to = 'cat') %>%
  group_by(var, cat) %>%
  count(var, cat) %>%
  arrange(match(var, var_order), match(cat, cat_order)) %>%
  ungroup() %>%
  mutate(
   pct_samp = round(((n / nrow(unweighted_input)) * 100), 1)
  ) %>%
  select(var, cat, n, pct_samp) %>% 
  full_join(region_census, by = "cat")

list_demos <- list(freq_demos_unweighted, gender_census, educ_census, 
                   ethnic_census, region_census)

freq_demos_comp <- list_demos %>% 
  reduce(left_join, by = "cat") %>% 
  filter(!(var %in% c("age_range", "clin_status"))) %>% 
  unite(census_count, c(Freq.x, Freq.y, Freq.x.x, Freq.y.y), sep = "", remove = T) %>% 
  mutate_at(vars(census_count), ~as.integer(str_replace_all(., "NA", ""))) %>% 
  mutate(census_pct = 100 * round(census_count/nrow(unweighted_input), 3),
         diff_pct = census_pct - pct_samp
         ) %>% 
  rename(input_count = n, input_pct = pct_samp) %>% 
  select(var, cat, input_count, census_count, input_pct, census_pct, diff_pct) %>% 
  arrange(desc(diff_pct))

rm(list = setdiff(ls(), ls(pattern = "input|comp|list")))

list_census <- list(list_demos[2:5],
     c("gender", "educ", "ethnic", "region")) %>%
  pmap(~ ..1 %>%
         rename_at(vars(cat), ~str_replace(., "cat", !!..2))
  )

names(list_census) <- c("gender_census", "educ_census", "ethnic_census", "region_census")

list2env(list_census, envir=.GlobalEnv)
```

<br>

#### Implement survey functions

We now use functions from the Survey package to first create a survey object based on the unweighted input data, then to rake that object against the census targets for `gender`, `educ`, `ethnic`, and `region`. In examining the resulting `rake_unweighted_input` object, I found the `prob` vector, which appears to contain the case-wise weights generated by the raking procedure. I labeled that vector `demot_wt`, bound it to the unweighted input data in the data frame`input_demo_wts`, and sorted that data frame by the `demo_wt` column.
```{r survey-funs, eval=FALSE}
unweighted_survey_object <- svydesign(ids = ~1, 
                                     data = unweighted_input, 
                                     weights = NULL)

rake_unweighted_input <- rake(design = unweighted_survey_object,
                          sample.margins = list(~gender, ~educ, ~ethnic, ~region),
                          population.margins = list(gender_census, educ_census, 
                                                    ethnic_census, region_census))

input_demo_wts <- bind_cols(
  rake_unweighted_input[["variables"]], 
  data.frame(rake_unweighted_input[["prob"]])
  ) %>% 
  rename(demo_wt = rake_unweighted_input...prob...) %>% 
  select(ID:clin_status, demo_wt, everything()) %>% 
  arrange(desc(demo_wt))

tail(input_demo_wts)

filter(input_demo_wts, between(demo_wt, .98, 1.02))

head(input_demo_wts)
```
The values in the `demo_wt` column at first glance appear to have the expected relationship with the demographic classifcations of their associated rows. For example, the `demo_wt` value of `0.163` associated with the the `female-no_HS_hispanic_northeast` classification is relatively distant from 1, as we can see from calling `tail(input_demo_wts)`.

When we examine the values of `demo_wt` clustered around 1 by calling `filter(input_demo_wts, between(demo_wt, .98, 1.02))`,  we see primarily the expected demographic categories of `HS_grad`, `black`, and `south`.

Here is where I run up against the limit of my understanding of the Survey functions and the output of `rake()`. Because the `female-no-HS-hispanic-northeast` cell is under-sampled relative to the census target, I would expect the `demo_wt` value to be _greater_ than 1, by the logic I described previously. That is, the impact of the existing cases in that cell on the sample statistics would need to be magnified by a multiplier greater than 1, to offset the fact that there are fewer cases in that cell than required by the census targets.

The fact that `demo_wt` is _less_ than 1 for the `female-no_HS-hispanic-northeast` cell is counter-intuitive to me. It seems like an inversion of what the value should be. This seeming inversion is present throughout the rows of `input_demo_wts`. Calling `head(input_demo_wts)` reveals that cases from the over-sampled crossing of `BA_plus` and `midwest` are associated with a `demo_wt` of `2.167`. This again is counter-intuitive; I would expect a `demo_wt` less than 1, to reduce the impact of these over-sampled categories on the sample statistics.

What to make of this? It seems that the values of the `prob` vector (my `demo_wt` column) are in fact describing the demographic distribution of the unweighted input sample. Specifically, it seems that these values represent the level of under- or oversampling- of each crossing of `gender`, `educ`, `ethnic`, and `region`, relative to the census targets for those crossings.

If I'm understanding this correctly, these `demo_wt` values cannot serve as demographic weighting multipliers in the manner that I described above in the __Problem Statement__ section. I suspect that the `demo_wt` values have a linear relationship with the multipliers I am seeking, but I'm unsure how to make the mathematical transformation. Or perhaps my desired multipliers are lurking in another function within the Survey package.

This is as far as I've taken it before reaching out to you for help. I'm grateful for your interest in helping with this problem. Thank you again, and please let me know if/when we get into a realm where it would make sense to convert this exchange into a paid consulting engagement. I'm sure we can work out payment and taxation.
